---
title: "Data Skills 05 - Plotting Time Series Data with ggplot2 - Complete"
author: "Christopher Prener, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output:
  github_document: default
  html_notebook: default
---

## Dependencies
This notebook requires two packages from the `tidyverse` as well as two additional packages:

```{r load-packages}
# tidyverse packages
library(dplyr)     # data wrangling
library(ggplot2)   # plotting
library(readr)     # read and write csv files

# manage file paths
library(here)      # manage file paths
```

## Load Data
For this session and our future sessions this spring, we'll focus on two sets of data - COVID cases and mortality from the New York Times and SARS-CoV-2 variants from the CDC. For today, we'll use pre-cleaned versions of both. First, we'll load the COVID case and mortality data:

```{r load-covid}
nyt_wide <- read_csv(here("data", "nyt_usa_covid_wide.csv"))
nyt_long <- read_csv(here("data", "nyt_usa_covid_long.csv"))
```

Next, you write a similar statement to load the CDC data:

```{r load-cdc}
cdc <- read_csv(here("data", "cdc_concentrations.csv"))
```

## Data Preparation
We have two versions of the New York Times data to look at - "wide" formatted data and "long" formatted data.

```{r}
head(nyt_wide, n = 10L)
```

Notice how we have one row per day, with columns representing different measures.

*As an aside, notice how I specified the `n = 10L`. The `L` confirms for `R` that we are passing an integer value. Usually, functions will work without it, but it is a good idea to be clear to `R` that you want an integer and only an integer. Ignoring this works until it doesn't!*

Next, we'll explore the long data:

```{r}
head(nyt_long, n = 40L)
```

Notice how we have one row per measure, with multiple rows per day. This is the format that we need to work in for plotting longitudinal data. Next week, we'll talk about the work needed to wrangle data into this format. For now, though, we'll move on to the plotting piece!

```{r clean-up}
rm(nyt_wide)
```

## Plotting with `geom_area`
`ggplot2` contains a number of tools for showing change over time. We'll start with `geom_area()` and review the main components of a `ggplot` call. Notice how we pipe our data into the `ggplot()` call below: we take our `nyt_long` data **then** subset to our selected measure **then** create our plot with those data.

```{r cases-area}
nyt_long %>%
  filter(measure == "cases_avg7_pc") %>%
  ggplot(mapping = aes(x = date, y = value)) +
    geom_area()
```

Recall from our last session that there are a number of ways we can augment our plots, including altering the `y` axis breaks, updating our labels, and adding a title and subtitle. We can also select a fill for our area data itself.

```{r cases-area-expanded}
nyt_long %>%
  filter(measure == "cases_avg7_pc") %>%
  ggplot(mapping = aes(x = date, y = value)) +
    geom_area(fill = "blue") +
    scale_y_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, by = .25)) +
    labs(
      title = "COVID Cases",
      subtitle = "United States",
      x = "Year",
      y = "Per capita Rate"
    )
```

One additional tool we can use is to alter our x axis to show months instead of years. We can do this with `scale_x_date()`. There are a number of key pieces:
  * Breaks can be specified in counts of days, weeks, months, or years
  * Labels require `scales::label_date()`
  * The labels can be a combination of values, like `%B` for the month name, `%b` for the month abbreviation, `%Y` for year, `%d` for day
  * Finally, we can experiment with the `expand` argument to force the x-axis labels outward - *this requires lots of trial and error!*
  * Placing the x-axis labels at an angle usually helps with readability using the `theme()` function's `axis.text.x` argument

```{r cases-area-complete}
nyt_long %>%
  filter(measure == "cases_avg7_pc") %>%
  ggplot(mapping = aes(x = date, y = value)) +
    geom_area(fill = "blue") +
    scale_x_date(breaks = "3 months", labels = scales::label_date("%b %Y"), expand = c(.1, .1)) + 
    scale_y_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, by = .25)) +
    labs(
      title = "COVID Cases",
      subtitle = "United States",
      x = "Year",
      y = "Per capita Rate"
    ) +
  theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Now, you experiment with `geom_area()` by using the `cdc` data to isolate `BA.5` observations when the region is equal to `USA`. Try to experiment with `x` axis breaks and labels to achieve different designs for your plot!

```{r variants-area}
cdc %>%
  filter(variant == "BA.5" & region == "USA") %>%
  ggplot(mapping = aes(x = week, y = pct)) +
    geom_area(fill = "orange") +
    scale_x_date(breaks = "2 months", labels = scales::label_date("%b %Y"), expand = c(.1,.1)) + 
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
    labs(
      title = "BA.5 as a Percentage of All Variants",
      subtitle = "United States",
      x = "Year",
      y = "Percentage"
    ) +
  theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

## Plotting with `geom_line`
An alternative to `geom_area()` is the `geom_line()` function, which works very similarly. One advantage of `geom_line()` from a design standpoint is "less ink" on the plot. This makes it easier to compare multiple trends. For example, we could leverage it to plot cases and deaths simultaneously. Most of the code here is the same as above, but I have altered the initial `filter()` statement to include deaths and also cleaned up those `measure` values so that they are more appropriate for plotting. To deal with two groups, notice that `color = measure` now appears in our aesthetic mapping instead of being hard-coded into `geom_line()`. I've also added `scale_color_brewer()` to tweak the line colors and the name of the legend.

```{r cases-deaths-line}
nyt_long %>%
  filter(measure %in% c("cases_avg7_pc", "deaths_avg7_pc") == TRUE) %>%
  mutate(measure = ifelse(measure == "cases_avg7_pc", "Cases", "Deaths")) %>%
  ggplot(mapping = aes(x = date, y = value, color = measure)) +
    geom_line() + 
    scale_x_date(breaks = "3 months", labels = scales::label_date("%b %Y")) + 
    scale_y_continuous(limits = c(0, 2.5), breaks = seq(0, 2.5, by = .25)) +
    scale_color_brewer(palette = "Set1", name = "Outcome") +
    labs(
      title = "COVID Cases and Deaths",
      subtitle = "United States",
      x = "Year",
      y = "Per capita Rate"
    ) +
  theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

This is often where people will jump in with a "dual y-axis" plot. `ggplot2` does not make it easy to build these, and honestly, that is a good thing. The dual y-axis plot can be a great way to mislead viewers. If you notice that your two distributions work on *very* different scales, step back and ask yourself if one plot is appropriate. There are other options out there!

Now, you take a turn - write out code to simultaneously plot BA.4 and BA.5 over time, coloring the lines for each sub-variant in a different color. Focus on the national data only!

```{r variants-line}
cdc %>%
  filter(region == "USA") %>%
  ggplot(mapping = aes(x = week, y = pct, color = variant)) +
    geom_line() + 
    scale_x_date(breaks = "2 months", labels = scales::label_date("%b %Y"), expand = c(.1,.1)) + 
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
    scale_color_brewer(palette = "Dark2", name = "Omicron\nSub-Variant") +
    labs(
      title = "BA.4 and BA.5 as a Percentage of All Variants",
      subtitle = "United States",
      x = "Year",
      y = "Percentage"
    ) +
  theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```

Finally, if you have time, create a plot of BA.5 over time between the US as a whole and HHS Region 1 (New England):

```{r region-line}
cdc %>%
  filter(variant == "BA.5") %>%
  mutate(region = ifelse(region == "HHS1", "New England", "National")) %>%
  ggplot(mapping = aes(x = week, y = pct, color = region)) +
    geom_line() + 
    scale_x_date(breaks = "2 months", labels = scales::label_date("%b %Y"), expand = c(.1,.1)) + 
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
    scale_color_brewer(palette = "Dark2", name = "Geography") +
    labs(
      title = "BA.5 as a Percentage of All Variants",
      subtitle = "United States and New England",
      x = "Year",
      y = "Percentage"
    ) +
  theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)
  )
```
